Towards Practical First-Order Model Counting

### 2025-01-20 doc/paper/paper.tex
Rework Example 8

\begin{example}[Counting bijections]\label{example:overall}
  Let us consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two domains $\Gamma$ and $\Delta$:
  \begin{equation}\label[formula]{eq:bijections}
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \end{equation}
  Let $\sigma$ be defined as $\sigma(\Gamma) \coloneqq \{\, 1, 2\,\}$, and
  $\sigma(\Delta) \coloneqq \{\,a, b\,\}$. Then \cref{eq:bijections} has two
  models:
  \[
    \{\, P(1, a), P(2, b), \neg P(1, b), \neg P(2, a) \,\} \qquad \text{and} \qquad \{\, P(1, b), P(2, a), \neg P(1, a), \neg P(2, b) \,\}.
  \]
\end{example}

@@

\subsubsection{Compilation}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item The algorithm’s recursive nature and its interaction with
          \textsc{Crane} could be explained more clearly
    \item More emphasis on the motivation for each part of \Cranetwo{} (e.g.,
          how \FindBaseCases addresses key issues in \textsc{Crane}) would add
          depth to the contribution
    \item this section should be expanded by 1 page (``a clear discussion of the
          innovation aspect and technical parts'')
  \end{itemize}
}

@@

\todo[inline]{Connect this with the preliminaries?}

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
the FOKC algorithm of the original \textsc{Crane} to compile the formula into
the three components: $\mathcal{E}$, $\mathcal{F}$, and $\mathcal{D}$. After
some algebraic simplification, $\mathcal{E}$ is passed to the \FindBaseCases
procedure (see \cref{sec:identifying}). For each base case $f(\mathbf{x})$, we
retrieve the logical formula $\mathcal{F}(f)$ associated with the function name
$f$ and simplify it using the \Propagate procedure (explained in detail in
\cref{sec:propagating}). We do this by iterating over all indices of
$\mathbf{x}$, where $x_{i}$ is a constant, and using \Propagate to simplify
$\psi$ by assuming that domain $\mathcal{D}(f, i)$ has size $x_{i}$. Finally, on
\autoref{line:final}, \CompileWithBaseCases recurses on these simplified
formulas and adds the resulting base case equations to $\mathcal{E}$.
\Cref{example:overall} below provides more detail.

@@

\todo[inline]{The example below builds on \cref{sec:crane}. Shorten it.}

\begin{example}\label{example:overalltwo}
  Let us examine the first solution returned by \Cranetwo{} for the
  bijection-counting problem in \cref{example:overall}. After
  \autoref{line:second}, we have
  \begin{align*}
    \mathcal{E} &= \left\{\,\begin{aligned}f(m, n) &= \sum_{l=0}^{n} \binom{n}{l}{(-1)}^{n-l}g(l, m),\\ g(l, m) &= g(l-1, m) + mg(l-1, m-1)\end{aligned}\,\right\};\\
    \mathcal{D} &= \{\, (f, 1) \mapsto \Gamma, (f, 2) \mapsto \Delta, (g, 1) \mapsto \Delta^{\top}, (g, 2) \mapsto \Gamma \,\},
  \end{align*}
  where $\Delta^{\top}$ is a new domain. (We omit the definition of
  $\mathcal{F}$ as the formulas can get a bit verbose.) Then \FindBaseCases
  identifies two base cases: $g(0, m)$ and $g(l, 0)$. In both cases,
  \CompileWithBaseCases recurses on the formula $\mathcal{F}(g)$ simplified by
  assuming that one of the domains is empty. In the first case, we recurse on
  the formula $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a
  predicate introduced by preprocessing with weights $w^{+}(S) = 1$ and
  $w^{-}(S) = -1$. Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case
  of $g(l, 0)$, \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula,
  resulting in $g(l, 0) = 1$.
\end{example}

It is worth noting that these base cases overlap when $l = m = 0$ but remain
consistent since $0^{0} = 1$. Generally, let $\phi$ be a formula with two
domains $\Gamma$ and $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the FOMC
of \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the
FOMC of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.

Finally, the main responsibility of the \Simplify procedure is to handle the
algebraic pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here: $n$ is a variable,
$a, b \in \mathbb{N}_{0}$ are constants, and $f$ is an expression that may
depend on $m$. \Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$. For instance, in the case of
\cref{example:overalltwo}, \Simplify transforms
$g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
$g(l, m) = g(l-1, m) + mg(l-1, m-1)$.

\todo[inline]{Check if the above is still clear given all the changes.}

@@

\todo[inline]{Prove correctness?}

@@

\begin{example}
  Consider the recursive function $g$ from \cref{example:overalltwo}.
  \FindBaseCases iterates over two function calls: $g(l-1, m)$ and
  $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the latter
  produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

@@

\begin{algorithm}[t]
  \caption{A sketch of the C++ program for the equations in \cref{example:overalltwo}, particularly highlighting the recursive definition of function $g$.}\label{alg:cpp}
  initialise $\Cache_{g(0, m)}$, $\Cache_{g(l, 0)}$, $\Cache_{g}$, and $\Cache_{f}$\;
  \DontPrintSemicolon
  \textbf{Function} $g_{0,m}(m)$\textbf{:} \dots\;
  \textbf{Function} $g_{l,0}(l)$\textbf{:} \dots\;
  \PrintSemicolon
  \Fn{$g(l, m)$}{
    \lIf{$(l, m) \in \Cache_{g}$}{\Return{$\Cache_{g}(l, m)$}}\label{line:cache}
    \lIf{$l = 0$}{\Return{$g_{0,m}(m)$}}\label{line:basecaseone}
    \lIf{$m = 0$}{\Return{$g_{l,0}(l)$}}\label{line:basecasetwo}
    $r \gets g(l-1, m) + mg(l-1,m-1)$\;
    $\Cache_{g}(l, m) \gets r$\;
    \Return{$r$}\;
  }
  \DontPrintSemicolon
  \textbf{Function} $f(m, n)$\textbf{:} \dots\;
  \PrintSemicolon
  \Fn{\Main}{
    $(m, n) \gets \Input{}$\;
    \Return{$f(m, n)$}\;
  }
\end{algorithm}

@@

We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by Dilkas and Belle~\cite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' Markov
logic network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In
\Ctwo{}, \FO{}, and \UFO{}, this problem can be formulated as
\[
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\]
or, equivalently, in conjunctive normal form as
\[
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\]
Finally, we include the bijection-counting problem previously utilised by Dilkas
and Belle~\cite{DBLP:conf/kr/DilkasB23}. Its formulation in \FO{} is described
in \cref{example:overall}. The equivalent formula in \Ctwo{} is
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{. }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{. }P(x, y)).
\]
Similarly, in \UFO{} the same formula can be written as
\[
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.

>>>>

\begin{example}[Counting bijections]\label{example:bijections}
  Let us consider the following formula (previously examined by Dilkas and
  Belle~\cite{DBLP:conf/kr/DilkasB23}) that defines predicate $P$ as a bijection
  between two domains $\Gamma$ and $\Delta$:
  \begin{equation}\label[formula]{eq:bijections}
    \begin{gathered}
      (\forall x \in \Gamma\text{. }\exists y \in \Delta\text{. }P(x, y))\land{}\\
      (\forall y \in \Delta\text{. }\exists x \in \Gamma\text{. }P(x, y))\land{}\\
      (\forall x \in \Gamma\text{. }\forall y, z \in \Delta\text{. }P(x, y) \land P(x, z) \Rightarrow y = z)\land{}\\
      (\forall x, z \in \Gamma\text{. }\forall y \in \Delta\text{. }P(x, y) \land P(z, y) \Rightarrow x = z).
    \end{gathered}
  \end{equation}
  Let $\sigma$ be defined as $\sigma(\Gamma) \coloneqq \{\, 1, 2\,\}$, and
  $\sigma(\Delta) \coloneqq \{\,a, b\,\}$. Then \cref{eq:bijections} has two
  models:
  \[
    \{\, P(1, a), P(2, b), \neg P(1, b), \neg P(2, a) \,\} \qquad \text{and} \qquad \{\, P(1, b), P(2, a), \neg P(1, a), \neg P(2, b) \,\}.
  \]
\end{example}

@@

\subsubsection{Compilation}\label{sec:compilation}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item More emphasis on the motivation for each part of \Cranetwo{} (e.g.,
          how \FindBaseCases addresses key issues in \textsc{Crane}) would add
          depth to the contribution
    \item this section should be expanded by 1 page (``a clear discussion of the
          innovation aspect and technical parts'')
  \end{itemize}
}

@@

\todo[inline,caption={}]{
  \begin{itemize}
    \item Reorganise this section (e.g., where do I put the description of
          Simplify?), bringing some TODOs from above.
    \item The algorithm’s recursive nature and its interaction with
          \textsc{Crane} could be explained more clearly
  \end{itemize}
}

\Cref{alg:compilewithbasecases} presents our overall approach for compiling a
formula into equations that include the necessary base cases. To begin, we use
\textsc{Crane} to compile the formula into the three components: $\mathcal{E}$,
$\mathcal{F}$, and $\mathcal{D}$ (as described in \cref{sec:compilation}). After
some algebraic simplification, $\mathcal{E}$ is passed to the \FindBaseCases
procedure (see \cref{sec:identifying}). For each base case $f(\mathbf{x})$, we
retrieve the logical formula $\mathcal{F}(f)$ associated with the function name
$f$ and simplify it using the \Propagate procedure (explained in detail in
\cref{sec:propagating}). We do this by iterating over all indices of
$\mathbf{x}$, where $x_{i}$ is a constant, and using \Propagate to simplify
$\psi$ by assuming that domain $\mathcal{D}(f, i)$ has size $x_{i}$. Finally, on
\autoref{line:final}, \CompileWithBaseCases recurses on these simplified
formulas and adds the resulting base case equations to $\mathcal{E}$.
\Cref{example:bijections} below provides more detail.

@@

\paragraph*{Simplify}
The main responsibility of the \Simplify procedure is to handle the algebraic
pattern $\sum_{m=0}^{n}[a \le m \le b] f(m)$. Here: $n$ is a variable,
$a, b \in \mathbb{N}_{0}$ are constants, and $f$ is an expression that may
depend on $m$. \Simplify transforms this pattern into
$f(a) + f(a+1) + \cdots + f(\min\{\, n, b \,\})$.

\begin{example}\label{example:bijectionstwo}
  Let us return to the bijection-counting problem from \cref{example:bijections}
  and its initial solution described in \cref{example:bijectionstwo}. On
  \autoref{line:second}, \Simplify transforms
  $g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)$ into
  $g(l, m) = g(l-1, m) + mg(l-1, m-1)$. Then \FindBaseCases identifies two base
  cases: $g(0, m)$ and $g(l, 0)$. In both cases, \CompileWithBaseCases recurses
  on the formula $\mathcal{F}(g)$ simplified by assuming that one of the domains
  is empty. In the first case, we recurse on the formula
  $\forall x \in \Gamma\text{. }S(x) \lor \neg S(x)$, where $S$ is a predicate
  introduced by preprocessing with weights $w^{+}(S) = 1$ and $w^{-}(S) = -1$.
  Hence, we obtain the base case $g(0, m) = 0^{m}$. In the case of $g(l, 0)$,
  \Propagate{$\psi$, $\Gamma$, $0$} returns an empty formula, resulting in
  $g(l, 0) = 1$.

  While these base cases overlap when $l = m = 0$, they remain consistent since
  $0^{0} = 1$. Generally, let $\phi$ be a formula with two domains $\Gamma$ and
  $\Delta$, and let $n, m \in \mathbb{N}_{0}$. Then the FOMC of
  \Propagate{$\phi$, $\Delta$, $n$} assuming $|\Gamma| = m$ is the same as the
  FOMC of \Propagate{$\phi$, $\Gamma$, $m$} assuming $|\Delta| = n$.
\end{example}

@@

@@

\begin{example}
  Consider the recursive function $g$ from \cref{example:bijectionstwo}.
  \FindBaseCases iterates over two function calls: $g(l-1, m)$ and
  $g(l-1, m-1)$. The former produces the base case $g(0, m)$, while the latter
  produces both $g(0, m)$ and $g(l, 0)$.
\end{example}

@@

\begin{algorithm}[t]
  \caption{A sketch of the C++ program for the equations in \cref{example:bijectionstwo}, particularly highlighting the recursive definition of function $g$.}\label{alg:cpp}
  initialise $\Cache_{g(0, m)}$, $\Cache_{g(l, 0)}$, $\Cache_{g}$, and $\Cache_{f}$\;
  \DontPrintSemicolon
  \textbf{Function} $g_{0,m}(m)$\textbf{:} \dots\;
  \textbf{Function} $g_{l,0}(l)$\textbf{:} \dots\;
  \PrintSemicolon
  \Fn{$g(l, m)$}{
    \lIf{$(l, m) \in \Cache_{g}$}{\Return{$\Cache_{g}(l, m)$}}\label{line:cache}
    \lIf{$l = 0$}{\Return{$g_{0,m}(m)$}}\label{line:basecaseone}
    \lIf{$m = 0$}{\Return{$g_{l,0}(l)$}}\label{line:basecasetwo}
    $r \gets g(l-1, m) + mg(l-1,m-1)$\;
    $\Cache_{g}(l, m) \gets r$\;
    \Return{$r$}\;
  }
  \DontPrintSemicolon
  \textbf{Function} $f(m, n)$\textbf{:} \dots\;
  \PrintSemicolon
  \Fn{\Main}{
    $(m, n) \gets \Input{}$\;
    \Return{$f(m, n)$}\;
  }
\end{algorithm}

@@

We compare these algorithms using three benchmarks from previous studies. The
first benchmark is the function-counting problem from \cref{example:functions},
previously examined by Dilkas and Belle~\cite{DBLP:conf/kr/DilkasB23}. The
second benchmark is a variant of the well-known `Friends and Smokers' Markov
logic network~\cite{DBLP:conf/aaai/SinglaD08,DBLP:conf/uai/BroeckCD12}. In
\Ctwo{}, \FO{}, and \UFO{}, this problem can be formulated as
\[
  (\forall x,y \in \Delta\text{. } S(x) \land F(x, y) \Rightarrow S(y)) \land (\forall x \in \Delta\text{. }S(x) \Rightarrow C(x))
\]
or, equivalently, in conjunctive normal form as
\[
  (\forall x,y \in \Delta\text{. }S(y) \lor \neg S(x) \lor \neg F(x, y)) \land (\forall x \in \Delta\text{. } C(x) \lor \neg S(x)).
\]
Finally, we include the bijection-counting problem previously utilised by Dilkas
and Belle~\cite{DBLP:conf/kr/DilkasB23}. Its formulation in \FO{} is described
in \cref{example:bijections}. The equivalent formula in \Ctwo{} is
\[
  (\forall x \in \Delta\text{. }\exists^{=1} y \in \Delta\text{. }P(x, y)) \land (\forall y \in \Delta\text{. }\exists^{=1} x \in \Delta\text{. }P(x, y)).
\]
Similarly, in \UFO{} the same formula can be written as
\[
  (\forall x, y \in \Delta\text{. }R(x) \lor \neg P(x, y)) \land (\forall x, y \in \Delta\text{. }S(x) \lor \neg P(y, x)) \land (|P| = |\Delta|),
\]
where $w^{-}(R) = w^{-}(S) = -1$.
